# Data import

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```

## Introduction

In this chapter, you'll learn how to read plain-text rectangular files into Julia. 
 Here, we'll only scratch the surface of data import, 
 but many of the principles will translate to other forms of data. 
 We'll finish with a few pointers to packages that are useful for other types of data.

### Prerequisites

In this chapter, you'll learn how to load flat files in Julia with a few different packages I called the "files" packages, which comes with Queryverse.  
Here is a list of them:

* [CSVFiles.jl](https://github.com/queryverse/CSVFiles.jl)

* [FeatherFile.jl](https://github.com/queryverse/FeatherFiles.jl)

* [ExcelFiles.jl](https://github.com/queryverse/ExcelFiles.jl)  

* [StatFiles.jl](https://github.com/queryverse/ParquetFiles.jl)  

* [ParquetFiles.jl](https://github.com/queryverse/ParquetFiles.jl)   

```{julia}
# If you haven't install Queryverse
#using Pkg
#Pkg.add("Queryverse")
# or add Queryverse in Pkg mode

# import Queryverse
using Queryverse, Dates, DataFrames, CSV
```

## Getting started

With R and Python, there are different functions for the vast majority of data formats. Some common ones are namely:
read_csv() for csv files, read_fwf() for fixed width files. Here is another great example of how Julia utilitizes multiple dispatch
to simplify code.  
  
With Julia, most data formats can be parsed and read with `load()`. The implementation of polymorphism allows Julia to parse the given
files according to its format.  
  
No more remembering different function names and and package names, `load()` is what you need!

```{julia}
heights = load("data/heights.csv") |> DataFrame
```

You can also supply an inline csv files. Instead of `load()` which takes in a file name, you can use a non-standard string to create tabular data.
Along with other types of file format.

```{julia}
using CSV;

data = CSV.read(IOBuffer("""a,b,c
                            1,2,3
                            4,5,6"""),
                DataFrame)

data = CSV.read(IOBuffer("""a;b;c
                            1;2;3
                            4;5;6"""),
                DataFrame)

# space-separated values
data = CSV.read(IOBuffer("""a b c
                            1 2 3
                            4 5 6"""), 
                DataFrame)

```



## Parsing a vector

Before we get into the details of how readr reads files from disk, 
we need to take a little detour to talk about the `parse()` functions. 
These functions take a character vector and return a more specialised vector like a logical, integer, or date:

```{julia}
parse.(Bool, ["true", "false"])
parse.(Int, ["1", "2", "3"])
parse.(Dates.Date, ["2010-01-01", "1979-10-14"])
```

These functions are useful in their own right, but are also an important building block. 
Once you've learned how the individual parsers work in this section, 
we'll circle back and see how they fit together to parse a complete file in the next section.

The `parse()` function takes a Datatype as the first argument and parse the string given in the second parameter.
```{julia}
# Bool being a datatype
typeof(Bool)

# Use tryparse to parse the string "true"
parse(Bool, "true")

# check the datatype of "true"
parse(Bool, "true") |> typeof

```

### `parse()` vs `tryparse()`

Both `parse()` and `tryparse()` works as a coverter for different datatypes. However, they react differently when they encounter
unparseable strings. `parse()` will throw an error if the string is cannot be convected, while `tryparse()` will still run but
return nothing, meaning nothing is convected in the case.
```{julia}
# nothing returns
tryparse(Int, "1.2")
```
### vector conversion
Knowing how to convert datatype for one string might not be as useful as knowing how to convert a vector of strings.
Using the build in vectorization feature of Julia, we can run the following code.
```{julia}
# converting a string
tryparse(Int, "1")

# Adding a dot/period after a function to vectorize
tryparse.(Int, ["1","2","3"])

# The vectorized function also works with missing/wrong values
tryparse.(Int, ["1","2","3", "", "not a number"])
```

Instead of returning nothing, the vectorized `tryparse()` now return a value of nothing. For the sake of the out understanding, we can
replace it with missing value.

```{julia}
n_list = tryparse.(Int, ["1","2","3", "", "not a number"]);
replace(n_list, nothing => missing)

```


Using parsers is mostly a matter of understanding what's available and how they deal with different types of input. 
There are eight particularly important parsers:

1.  `parse_logical()` and `parse_integer()` parse logicals and integers
    respectively. There's basically nothing that can go wrong with these
    parsers so I won't describe them here further.
    
1.  `parse_double()` is a strict numeric parser, and `parse_number()` 
    is a flexible numeric parser. These are more complicated than you might
    expect because different parts of the world write numbers in different
    ways.
    
1.  `parse_character()` seems so simple that it shouldn't be necessary. But
    one complication makes it quite important: character encodings.

1.  `parse_factor()` create factors, the data structure that R uses to represent
    categorical variables with fixed and known values.

1.  `parse_datetime()`, `parse_date()`, and `parse_time()` allow you to
    parse various date & time specifications. These are the most complicated
    because there are so many different ways of writing dates.

The following sections describe these parsers in more detail.

### Numbers

It seems like it should be straightforward to parse a number, but three problems make it tricky:

1. People write numbers differently in different parts of the world.
   For example, some countries use `.` in between the integer and fractional 
   parts of a real number, while others use `,`.
   
1. Numbers are often surrounded by other characters that provide some
   context, like "$1000" or "10%".

1. Numbers often contain "grouping" characters to make them easier to read, 
   like "1,000,000", and these grouping characters vary around the world.

These problems could be solved by regular expressions along with `replace()`.  
For example
```{julia}
replace.(["10%", "50%", "100%"], r"%" => "")
```
### Strings {#readr-strings}

It seems like `parse_character()` should be really simple --- it could just return its input. 
Unfortunately life isn't so simple, as there are multiple ways to represent the same string. 
To understand what's going on, we need to dive into the details of how computers represent strings. 
In Julia, we can get at the underlying representation of a string using a b string:

```{julia}
b"string"

String(b"string")
```

Each hexadecimal number represents a byte of information: `48` is H, `61` is a, and so on. The mapping from hexadecimal number to character is called the encoding, and in this case the encoding is called ASCII. ASCII does a great job of representing English characters, because it's the __American__ Standard Code for Information Interchange.

Things get more complicated for languages other than English. In the early days of computing there were many competing standards for encoding non-English characters, and to correctly interpret a string you needed to know both the values and the encoding. For example, two common encodings are Latin1 (aka ISO-8859-1, used for Western European languages) and Latin2 (aka ISO-8859-2, used for Eastern European languages). In Latin1, the byte `b1` is "±", but in Latin2, it's "ą"! Fortunately, today there is one standard that is supported almost everywhere: UTF-8. UTF-8 can encode just about every character used by humans today, as well as many extra symbols (like emoji!).

Julia uses UTF-8 everywhere: it assumes your data is UTF-8 encoded when you read it, and always uses it when writing. This is a good default, but will fail for data produced by older systems that don't understand UTF-8. If this happens to you, your strings will look weird when you print them. Sometimes just one or two characters might be messed up; other times you'll get complete gibberish.

To fix the problem you need to the package [StringEncodings](https://github.com/JuliaStrings/StringEncodings.jl)

### Factors {#readr-factors}

R uses factors to represent categorical variables that have a known set of possible values. Give `parse_factor()` a vector of known `levels` to generate a warning whenever an unexpected value is present:

```{r}
fruit <- c("apple", "banana")
parse_factor(c("apple", "banana", "bananana"), levels = fruit)
```

But if you have many problematic entries, it's often easier to leave as character vectors and then use the tools you'll learn about in [strings] and [factors] to clean them up.

### Dates, date-times, and times {#readr-datetimes}

You pick between three parsers depending on whether you want a date (the number of days since 1970-01-01), a date-time (the number of seconds since midnight 1970-01-01), or a time (the number of seconds since midnight). When called without any additional arguments:

*   `parse_datetime()` expects an ISO8601 date-time. ISO8601 is an
    international standard in which the components of a date are
    organised from biggest to smallest: year, month, day, hour, minute, 
    second.
    
    ```{r}
    parse_datetime("2010-10-01T2010")
    # If time is omitted, it will be set to midnight
    parse_datetime("20101010")
    ```
    
    This is the most important date/time standard, and if you work with
    dates and times frequently, I recommend reading
    <https://en.wikipedia.org/wiki/ISO_8601>
    
*   `parse_date()` expects a four digit year, a `-` or `/`, the month, a `-` 
    or `/`, then the day:
    
    ```{r}
    parse_date("2010-10-01")
    ```

*   `parse_time()` expects the hour, `:`, minutes, optionally `:` and seconds, 
    and an optional am/pm specifier:
  
    ```{r}
    library(hms)
    parse_time("01:10 am")
    parse_time("20:10:01")
    ```
    
    Base R doesn't have a great built in class for time data, so we use 
    the one provided in the hms package.

If these defaults don't work for your data you can supply your own date-time `format`, built up of the following pieces:

Year
: `%Y` (4 digits). 
: `%y` (2 digits); 00-69 -> 2000-2069, 70-99 -> 1970-1999.

Month
: `%m` (2 digits).
: `%b` (abbreviated name, like "Jan").
: `%B` (full name, "January").

Day
: `%d` (2 digits).
: `%e` (optional leading space).

Time
: `%H` 0-23 hour.
: `%I` 0-12, must be used with `%p`.
: `%p` AM/PM indicator.
: `%M` minutes.
: `%S` integer seconds.
: `%OS` real seconds. 
: `%Z` Time zone (as name, e.g. `America/Chicago`). Beware of abbreviations:
  if you're American, note that "EST" is a Canadian time zone that does not
  have daylight savings time. It is _not_ Eastern Standard Time! We'll
  come back to this [time zones].
: `%z` (as offset from UTC, e.g. `+0800`). 

Non-digits
: `%.` skips one non-digit character.
: `%*` skips any number of non-digits.

The best way to figure out the correct format is to create a few examples in a character vector, and test with one of the parsing functions. For example:

```{r}
parse_date("01/02/15", "%m/%d/%y")
parse_date("01/02/15", "%d/%m/%y")
parse_date("01/02/15", "%y/%m/%d")
```

If you're using `%b` or `%B` with non-English month names, you'll need to set the  `lang` argument to `locale()`. See the list of built-in languages in `date_names_langs()`, or if your language is not already included, create your own with `date_names()`.

```{r}
parse_date("1 janvier 2015", "%d %B %Y", locale = locale("fr"))
```

## Parsing a file

Now that you've learned how to parse an individual vector, it's time to return to the beginning and explore how readr parses a file. There are two new things that you'll learn about in this section:

1. How readr automatically guesses the type of each column.
1. How to override the default specification.

### Strategy

readr uses a heuristic to figure out the type of each column: it reads the first 1000 rows and uses some (moderately conservative) heuristics to figure out the type of each column. You can emulate this process with a character vector using `guess_parser()`, which returns readr's best guess, and `parse_guess()` which uses that guess to parse the column:

```{r}
guess_parser("2010-10-01")
guess_parser("15:01")
guess_parser(c("TRUE", "FALSE"))
guess_parser(c("1", "5", "9"))
guess_parser(c("12,352,561"))

str(parse_guess("2010-10-10"))
```

The heuristic tries each of the following types, stopping when it finds a match:

* logical: contains only "F", "T", "FALSE", or "TRUE".
* integer: contains only numeric characters (and `-`).
* double: contains only valid doubles (including numbers like `4.5e-5`).
* number: contains valid doubles with the grouping mark inside.
* time: matches the default `time_format`.
* date: matches the default `date_format`.
* date-time: any ISO8601 date.

If none of these rules apply, then the column will stay as a vector of strings.

### Problems

These defaults don't always work for larger files. There are two basic problems:

1.  The first thousand rows might be a special case, and readr guesses
    a type that is not sufficiently general. For example, you might have 
    a column of doubles that only contains integers in the first 1000 rows. 

1.  The column might contain a lot of missing values. If the first 1000
    rows contain only `NA`s, readr will guess that it's a logical 
    vector, whereas you probably want to parse it as something more
    specific.

readr contains a challenging CSV that illustrates both of these problems:

```{r}
challenge <- read_csv(readr_example("challenge.csv"))
```

(Note the use of `readr_example()` which finds the path to one of the files included with the package)

There are two printed outputs: the column specification generated by looking at the first 1000 rows, and the first five parsing failures. It's always a good idea to explicitly pull out the `problems()`, so you can explore them in more depth:

```{r}
problems(challenge)
```

A good strategy is to work column by column until there are no problems remaining. Here we can see that there are a lot of parsing problems with the `y` column. If we look at the last few rows, you'll see that they're dates stored in a character vector:  

```{r}
tail(challenge)
```

That suggests we need to use a date parser instead. To fix the call, start by copying and pasting the column specification into your original call:

```{r, eval = FALSE}
challenge <- read_csv(
  readr_example("challenge.csv"), 
  col_types = cols(
    x = col_double(),
    y = col_logical()
  )
)
```

Then you can fix the type of the `y` column by specifying that `y` is a date column:

```{r}
challenge <- read_csv(
  readr_example("challenge.csv"), 
  col_types = cols(
    x = col_double(),
    y = col_date()
  )
)
tail(challenge)
```

Every `parse_xyz()` function has a corresponding `col_xyz()` function. You use `parse_xyz()` when the data is in a character vector in R already; you use `col_xyz()` when you want to tell readr how to load the data.

I highly recommend always supplying `col_types`, building up from the print-out provided by readr. This ensures that you have a consistent and reproducible data import script. If you rely on the default guesses and your data changes, readr will continue to read it in. If you want to be really strict, use `stop_for_problems()`: that will throw an error and stop your script if there are any parsing problems.

### Other strategies

There are a few other general strategies to help you parse files:

*   In the previous example, we just got unlucky: if we look at just
    one more row than the default, we can correctly parse in one shot:
   
    ```{r}
    challenge2 <- read_csv(readr_example("challenge.csv"), guess_max = 1001)
    challenge2
    ```

*   Sometimes it's easier to diagnose problems if you just read in all
    the columns as character vectors:
   
    ```{r}
    challenge2 <- read_csv(readr_example("challenge.csv"), 
      col_types = cols(.default = col_character())
    )
    ```
    
    This is particularly useful in conjunction with `type_convert()`,
    which applies the parsing heuristics to the character columns in a data
    frame.

    ```{r}
    df <- tribble(
      ~x,  ~y,
      "1", "1.21",
      "2", "2.32",
      "3", "4.56"
    )
    df
    
    # Note the column types
    type_convert(df)
    ```
    
*   If you're reading a very large file, you might want to set `n_max` to
    a smallish number like 10,000 or 100,000. That will accelerate your 
    iterations while you eliminate common problems.

*   If you're having major parsing problems, sometimes it's easier
    to just read into a character vector of lines with `read_lines()`,
    or even a character vector of length 1 with `read_file()`. Then you
    can use the string parsing skills you'll learn later to parse
    more exotic formats.

## Writing to a file

readr also comes with two useful functions for writing data back to disk: `write_csv()` and `write_tsv()`. Both functions increase the chances of the output file being read back in correctly by:

* Always encoding strings in UTF-8.
  
* Saving dates and date-times in ISO8601 format so they are easily
  parsed elsewhere.

If you want to export a csv file to Excel, use `write_excel_csv()` --- this writes a special character (a "byte order mark") at the start of the file which tells Excel that you're using the UTF-8 encoding.

The most important arguments are `x` (the data frame to save), and `path` (the location to save it). You can also specify how missing values are written with `na`, and if you want to `append` to an existing file.

```{r, eval = FALSE}
write_csv(challenge, "challenge.csv")
```

Note that the type information is lost when you save to csv:

```{r, warning = FALSE}
challenge
write_csv(challenge, "challenge-2.csv")
read_csv("challenge-2.csv")
```

This makes CSVs a little unreliable for caching interim results---you need to recreate the column specification every time you load in. There are two alternatives:

1.  `write_rds()` and `read_rds()` are uniform wrappers around the base 
    functions `readRDS()` and `saveRDS()`. These store data in R's custom 
    binary format called RDS:
    
    ```{r}
    write_rds(challenge, "challenge.rds")
    read_rds("challenge.rds")
    ```
  
1.  The feather package implements a fast binary file format that can
    be shared across programming languages:
    
    ```{r, eval = FALSE}
    library(feather)
    write_feather(challenge, "challenge.feather")
    read_feather("challenge.feather")
    #> # A tibble: 2,000 x 2
    #>       x      y
    #>   <dbl> <date>
    #> 1   404   <NA>
    #> 2  4172   <NA>
    #> 3  3004   <NA>
    #> 4   787   <NA>
    #> 5    37   <NA>
    #> 6  2332   <NA>
    #> # ... with 1,994 more rows
    ```

Feather tends to be faster than RDS and is usable outside of R. RDS supports list-columns (which you'll learn about in [many models]); feather currently does not.

```{r, include = FALSE}
file.remove("challenge-2.csv")
file.remove("challenge.rds")
```

## Other types of data

To get other types of data into R, we recommend starting with the tidyverse packages listed below. They're certainly not perfect, but they are a good place to start. For rectangular data:

* __haven__ reads SPSS, Stata, and SAS files.

* __readxl__ reads excel files (both `.xls` and `.xlsx`).

* __DBI__, along with a database specific backend (e.g. __RMySQL__, 
  __RSQLite__, __RPostgreSQL__ etc) allows you to run SQL queries against a 
  database and return a data frame.

For hierarchical data: use __jsonlite__ (by Jeroen Ooms) for json, and __xml2__ for XML. Jenny Bryan has some excellent worked examples at <https://jennybc.github.io/purrr-tutorial/>.

For other file types, try the [R data import/export manual](https://cran.r-project.org/doc/manuals/r-release/R-data.html) and the [__rio__](https://github.com/leeper/rio) package.


## Completing with R and Python
Using multithreading, Julia outperforms python in reading data and is as fast as R's fread.
Here an [article](https://towardsdatascience.com/the-great-csv-showdown-julia-vs-python-vs-r-aa77376fb96).
