# DataTable

## Introduction

Throughout this book we work with "DataTable" instead of Julia's `dataframe`. Just like the more popular DataFrames, DataTables _are_ data frames, we use DataTable in this book since it is the default table formet used in Queryverse. However, you are more than welcome to use DataFrame.jl , IndexedTable.jl, and TypedTables.jl as alternatives. In most places, I'll use the term DataTAble and data frame interchangeably; when I want to draw particular attention to other data packages, I'll call them their names. 

If this chapter leaves you wanting to learn more about DataTable, you might enjoy !(https://juliadata.github.io/DataTables.jl/stable/man/getting_started.html).

### Prerequisites

In this chapter we'll explore the __DataTables__ package, part of the core queryverse.

```{julia}
# If you haven't install Queryverse and RDatasets
#using Pkg
#Pkg.add("Queryverse")
#Pkg.add("Rdatasets")
# or add Queryverse, RDatasets in Pkg mode

# import queryverse
using RDatasets, Queryverse
```

## Creating DataTable

Almost all of the functions that you'll use in this book are based on DataTable as their data source, as DataTable are one of the unifying features of the queryverse. When importing data, you might want to convert data into a DataTable. You can do that with `DataTable()`:

```{julia}
iris = dataset("datasets", "iris")
dtiris = DataTable(iris)
```

You can create a new tibble from individual vectors with `tibble()`. `tibble()` will automatically recycle inputs of length 1, and allows you to refer to variables that you just created, as shown below.

```{r}
tibble(
  x = 1:5, 
  y = 1, 
  z = x ^ 2 + y
)
```

If you're already familiar with `data.frame()`, note that `tibble()` does much less: it never changes the type of the inputs (e.g. it never converts strings to factors!), it never changes the names of variables, and it never creates row names.

It's possible for a tibble to have column names that are not valid R variable names, aka __non-syntactic__ names. For example, they might not start with a letter, or they might contain unusual characters like a space. To refer to these variables, you need to surround them with backticks, `` ` ``:

```{r}
tb <- tibble(
  `:)` = "smile", 
  ` ` = "space",
  `2000` = "number"
)
tb
```

You'll also need the backticks when working with these variables in other packages, like ggplot2, dplyr, and tidyr.

Another way to create a tibble is with `tribble()`, short for **tr**ansposed tibble.  `tribble()` is customised for data entry in code: column headings are defined by formulas (i.e. they start with `~`), and entries are separated by commas. This makes it possible to lay out small amounts of data in easy to read form.

```{r}
tribble(
  ~x, ~y, ~z,
  #--|--|----
  "a", 2, 3.6,
  "b", 1, 8.5
)
```

I often add a comment (the line starting with `#`), to make it really clear where the header is.

## Tibbles vs. data.frame

There are two main differences in the usage of a tibble vs. a classic `data.frame`: printing and subsetting.

### Printing

Tibbles have a refined print method that shows only the first 10 rows, and all the columns that fit on screen. This makes it much easier to work with large data. In addition to its name, each column reports its type, a nice feature borrowed from `str()`:

```{r}
tibble(
  a = lubridate::now() + runif(1e3) * 86400,
  b = lubridate::today() + runif(1e3) * 30,
  c = 1:1e3,
  d = runif(1e3),
  e = sample(letters, 1e3, replace = TRUE)
)
```

Tibbles are designed so that you don't accidentally overwhelm your console when you print large data frames. But sometimes you need more output than the default display. There are a few options that can help.

First, you can explicitly `print()` the data frame and control the number of rows (`n`) and the `width` of the display. `width = Inf` will display all columns:

```{r, eval = FALSE}
nycflights13::flights %>% 
  print(n = 10, width = Inf)
```

You can also control the default print behaviour by setting options:

* `options(tibble.print_max = n, tibble.print_min = m)`: if more than `n`
  rows, print only `m` rows. Use `options(tibble.print_min = Inf)` to always
  show all rows.

* Use `options(tibble.width = Inf)` to always print all columns, regardless
  of the width of the screen.

You can see a complete list of options by looking at the package help with `package?tibble`.

A final option is to use RStudio's built-in data viewer to get a scrollable view of the complete dataset. This is also often useful at the end of a long chain of manipulations.

```{r, eval = FALSE}
nycflights13::flights %>% 
  View()
```

### Subsetting

So far all the tools you've learned have worked with complete data frames. If you want to pull out a single variable, you need some new tools, `$` and `[[`. `[[` can extract by name or position; `$` only extracts by name but is a little less typing.

```{r}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

# Extract by name
df$x
df[["x"]]

# Extract by position
df[[1]]
```

To use these in a pipe, you'll need to use the special placeholder `.`:

```{r}
df %>% .$x
df %>% .[["x"]]
```

Compared to a `data.frame`, tibbles are more strict: they never do partial matching, and they will generate a warning if the column you are trying to access does not exist.

## Interacting with older code

Some older functions don't work with tibbles. If you encounter one of these functions, use `as.data.frame()` to turn a tibble back to a `data.frame`:

```{r}
class(as.data.frame(tb))
```

The main reason that some older functions don't work with tibble is the `[` function.  We don't use `[` much in this book because `dplyr::filter()` and `dplyr::select()` allow you to solve the same problems with clearer code (but you will learn a little about it in [vector subsetting](#vector-subsetting)). With base R data frames, `[` sometimes returns a data frame, and sometimes returns a vector. With tibbles, `[` always returns another tibble.

## Exercises

1.  How can you tell if an object is a tibble? (Hint: try printing `mtcars`,
    which is a regular data frame). 

1.  Compare and contrast the following operations on a `data.frame` and 
    equivalent tibble. What is different? Why might the default data frame
    behaviours cause you frustration?
    
    ```{r, eval = FALSE}
    df <- data.frame(abc = 1, xyz = "a")
    df$x
    df[, "xyz"]
    df[, c("abc", "xyz")]
    ```

1.  If you have the name of a variable stored in an object, e.g. `var <- "mpg"`,
    how can you extract the reference variable from a tibble?

1.  Practice referring to non-syntactic names in the following data frame by:

    1.  Extracting the variable called `1`.

    1.  Plotting a scatterplot of `1` vs `2`.

    1.  Creating a new column called `3` which is `2` divided by `1`.
        
    1.  Renaming the columns to `one`, `two` and `three`. 
    
    ```{r}
    annoying <- tibble(
      `1` = 1:10,
      `2` = `1` * 2 + rnorm(length(`1`))
    )
    ```

1.  What does `tibble::enframe()` do? When might you use it?

1.  What option controls how many additional column names are printed
    at the footer of a tibble?
